### 2014-04-07 text/Chapter03.tex
Put all coeffect examples in chapter 3

Software developers as well as programming language researchers choose abstractions based not 
just on how appropriate they are. Other factors may include social aspects -- how well is the 
abstraction known, how well is it documented and whether it is a standard tool of the 
\emph{research programme}\footnote{ Research programme, as introduced by Lakatos \cite{philosophy-lakatos},
is a network of scientists sharing the same basic assumptions and techniques.}.

In Chapter~\ref{ch:intro}, we argued that context-awareness had, so far, only limited influence 
on the design of programming languages, possibly because it is a challenge that is not easy to
see. However, many of the properties that we treat uniformly as \emph{coeffects} have been
previously tracked by other means. This includes special-purpose type systems, systematic 
approaches arising from modal logic S4, as well as techniques based on abstractions designed 
for other purpose, most frequently monads.

@@

The difference between effects and coeffects becomes apparent when we consider lambda abstraction.
The typical lambda abstraction rule for effect systems looks as (\emph{abs-eff}) in 
Figure~\ref{fig:applications-abs}. Wadler and Thiemann~\cite{monads-effects-marriage} explain how 
the effect analysis works `as follows:
%
\begin{quote}
\emph{In the rule for abstraction, the effect is empty because evaluation immediately
returns the function, with no side effects. The effect on the function arrow
is the same as the effect for the function body, because applying the function will
have the same side effects as evaluating the body.}
\end{quote}
%
This is the key property of \emph{output impurity}. The effects are only produced when the
function is evaluated and so the effects of the body are attached to the function. A recent
work by Tate~\cite{effects-producer-semantics} uses the term \emph{producer} effect systems
for such standard systems and characterises them as follows:
%
\begin{quote}
\emph{Indeed, we will define an effect as a producer effect if all computations with that 
effect can be thunked as ``pure'' computations for a domain-specific notion of purity.}
\end{quote} 
%
The thunking is typically a lambda abstraction -- given an effectful expression $e$, the
function $\lambda x.e$ is effect free value (thunk) that delays all effects.

Coeffects do not follow this pattern. Indeed, we can see this even when we consider the classical
notion of context. Standard lambda abstraction (\emph{app-pure}) in Figure~\ref{fig:applications-abs}
splits the free-variable context of an expression between variables defined at the \emph{declaration 
site} (the lexical scope) and those provided by the \emph{call site} (bound variable, or the 
parameter). This is also the case for other notions of context that we consider in this thesis.

@@

Aside from the standard free-variable context, we consider two notions of context -- 
\emph{flat} refers to overall properties of the environment and \emph{structural} refer to 
properties attached to individual variables. Similarly to free-variable context, both of these
notions split the context between declaration-site (lexical scope) and call-site (dynamic scope).

@@

@@

>>>>

Software developers as well as programming language researchers choose abstractions based not 
just on how appropriate they are. Other factors may include social aspects -- how well is the 
abstraction known, how well is it documented and whether it is a standard tool of the 
\emph{research programme}\footnote{ Research programme, as introduced by Lakatos \cite{philosophy-lakatos},
is a network of scientists sharing the same basic assumptions and techniques.}. This may
partly be why no unified context tracking mechanism has been developed so far.

In Chapter~\ref{ch:intro}, we argued that context-awareness had, so far, only limited influence 
on the design of programming languages because it is a challenge that is not easy to
see. However, many of the properties that this thesis treats uniformly as \emph{coeffects} have been
previously tracked by other means. This includes special-purpose type systems, systematic 
approaches arising from modal logic S4, as well as techniques based on abstractions designed 
for other purpose, most frequently monads.

@@

The difference between effects and coeffects becomes apparent when we consider lambda abstraction.
The typical lambda abstraction rule for effect systems looks as (\emph{abs-eff}) in 
Figure~\ref{fig:applications-abs}. Wadler and Thiemann~\cite{monads-effects-marriage} explain how 
the effect analysis works as follows:
%
\begin{quote}
\emph{In the rule for abstraction, the effect is empty because evaluation immediately
returns the function, with no side effects. The effect on the function arrow
is the same as the effect for the function body, because applying the function will
have the same side effects as evaluating the~body.}
\end{quote}
%
This is the key property of \emph{output impurity}. The effects are only produced when the
function is evaluated and so the effects of the body are attached to the function. A recent
work by Tate~\cite{effects-producer-semantics} uses the term \emph{producer} effect systems
for such standard systems and characterises them as follows:
%
\begin{quote}
\emph{Indeed, we will define an effect as a producer effect if all computations with that 
effect can be thunked as ``pure'' computations for a domain-specific notion of purity.}
\end{quote} 
%
The thunking is typically performed by a lambda abstraction -- given an effectful expression 
$e$, the function $\lambda x.e$ is an effect free value (thunk) that delays all effects.
As shown in the next section, contextual properties do not follow this pattern.

@@

We look at three notions of context. The first is the standard free-variable context in 
$\lambda$-calculus. This is well understood and we use it to demonstrate how contextual 
properties behave. Then we consider two notions of context introduced in this thesis --
\emph{flat coeffects} refer to overall properties of the environment and \emph{structural coeffects} 
refer to properties attached to individual variables.

\paragraph{Variable coeffects.}

In standard $\lambda$-calculus, variable access can be seen as a primitive operation that 
accesses the context. The expression $x$ introduces a context requirement -- the expression 
is typeable only in a context that contains $x:\tau$ for some type $\tau$.

The standard lambda abstraction (\emph{abs-pure}), shown in Figure~\ref{fig:applications-abs},
splits the free-variable context of an expression into two parts. The value of the parameter
has to be provided by the \emph{call site} (dynamic scope) and the remaining values are provided 
by the \emph{declaration site} (lexical scope). Here, the splitting is determined syntactically --
the notation $\lambda x.e$ names the variable whose value comes from the call site.

The flat and structural coeffects behave in the same way. They also split context-requirements
between the declaration site and call site, but they do it in two different ways.

@@

In Chapter~\ref{ch:intro}, we used \emph{resources} in a distributed system as an example of flat
coeffects. These could be, for example, a database, GPS sensor or access to current time.
We also outlined that such context requirements can be tracked as part of the typing assumption,
for example, say we have an expression $e$ that requires GPS and current time. The context of
such expression will be $\coctx{\Gamma}{ \aclrd{\ident{\{ gps, time \}}} }$.

The interesting case is when we construct a lambda function $\lambda x.e$, marshall it and
send it to another node. In that case, the context requirements can be satisfied in a number of
ways. When the same resource is available at the target machine (\eg.~current time), we can
send the function with a context requirement and \emph{rebind} the resource. However, if the
resource is not available (\eg.~GPS on the server), we need to capture a \emph{remote reference}.

In the example discussed here, $\lambda x.e$ would require a context 
$\coctx{\Gamma}{ \aclrd{\ident{\{ gps \}}} }$

Context requirements in distributed systems can be satisfied in a number of ways. When the 
resource is available locally,

@@

We discuss this example in detail in Section~X.
